---
title: "10 - Analyse Data Sub-population"
author: "Mick Cooney"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    number_sections: true
    fig_caption: yes
    theme: cerulean
  pdf_document: default
---

<!--
(Title: 10 - Analyse Data Sub-population
Author: Mick Cooney
Date: `r Sys.Date()`
Abstract: The abstract for this article goes here
Keywords:
-->


```{r knit_opts, include=FALSE}
rm(list = ls())

knitr::opts_chunk$set(tidy       = FALSE
                     ,cache      = FALSE
                     ,fig.height =     8
                     ,fig.width  =    11
                     )

library(tidyverse)
library(forcats)
library(scales)
library(lubridate)
library(boot)


options(width = 90L)

set.seed(42)
```


# Introduction

This workbook loads the lifebook data and creates a sample subpopulation of the data to be analysed in the rest of the project.

```{r load_data, echo=TRUE}
lifebook_tbl <- read_csv("data/dataset.csv", progress = FALSE) %>%
    mutate_if(is.character, as.factor)

glimpse(lifebook_tbl)
```

Having loaded the data, we now separate the data into four categories: logical, numeric, categorical, and date/time.

```{r determine_column_datatypes, echo=TRUE}
drop_var <- c('policy_id', 'edname', 'sa_id')
subpop_var <- 'subpop'

subpop_vals <- lifebook_tbl[[subpop_var]] == TRUE
subpop_tbl  <- lifebook_tbl %>% filter(subpop_vals)

datatypes_tbl <- lifebook_tbl %>%
    select(-one_of(c(subpop_var, drop_var))) %>%
    summarise_each(funs(class)) %>%
    gather('variable','datatype')

var_types <- datatypes_tbl %>%
    .[['datatype']] %>%
    unique

gen_list <- lapply(var_types
                   ,function(x) datatypes_tbl %>% filter(datatype == x) %>% .[['variable']])

names(gen_list) <- var_types

type_list <- list(
    categorical = c(gen_list$factor, gen_list$logical)
   ,numeric     = c(gen_list$numeric, gen_list$integer)
   ,datetime    = gen_list$Date
)
```


# Initial Visualisations

We first create some simple plots based on the indicator.

```{r count_plot, echo=TRUE}
ggplot(lifebook_tbl) +
    geom_bar(aes(x = subpop)) +
    scale_y_continuous(labels = comma) +
    ylab("Count")
```


# Facet Plots

```{r create_faceting_formula, echo=TRUE}
facet_formula <- formula(paste0("~", subpop_var))
```

## Categorical Variables

```{r create_categorical_facetplots, echo=TRUE}
for(plot_var in type_list$categorical) {
    cat(paste0("Plot Var: ", plot_var, "\n"))
    
    var_plot <- ggplot(lifebook_tbl) +
        geom_bar(aes_string(x = plot_var)) +
        facet_wrap(facet_formula, scales = 'free') +
        scale_y_continuous(labels = comma) +
        xlab(plot_var) +
        ylab("Count") +
        theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))

    print(var_plot)
}
```

## Numeric Variables

```{r create_numerical_facetplots, echo=TRUE}
for(plot_var in type_list$numeric) {
    cat(paste0("Plot Var: ", plot_var, "\n"))
    
    var_plot <- ggplot(lifebook_tbl) +
        geom_histogram(aes_string(x = plot_var), bins = 50) +
        facet_wrap(facet_formula, scales = 'free') +
        scale_x_continuous(labels = comma) +
        scale_y_continuous(labels = comma) +
        xlab(plot_var) +
        ylab("Count")

    print(var_plot)
}
```

## Date/Time Variables

```{r create_datetime_facetplots, echo=TRUE}
for(plot_var in type_list$datetime) {
    cat(paste0("Plot Var: ", plot_var, "\n"))
    
    var_plot <- ggplot(lifebook_tbl) +
        geom_histogram(aes_string(x = plot_var), bins = 50) +
        facet_wrap(facet_formula, scales = 'free') +
        scale_x_date(date_labels = '%Y-%m-%d') +
        scale_y_continuous(labels = comma) +
        xlab(plot_var) +
        ylab("Count") +
        theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
        

    print(var_plot)
}
```


# Bootstrap Checks

This approach compares the data in the subpopulation against a number of bootstrap samples from the full dataset, and then plots the two against each other.

```{r setup_bootstrap, echo=TRUE}
bootstrap_count <- 250

calc_bootstrap_stats <- function(x, b) {
    if(missing(b)) b <- seq_along(x)
    
    use_x     <- x[b]
    samp_mean <- mean(use_x)
    samp_perc <- quantile(use_x
                         ,type = 1
                         ,probs = c(0.50, 0.01, 0.10, 0.25, 0.75, 0.90, 0.99))

    return(c(mean = samp_mean, samp_perc))    
}

generate_bootstrap_props <- function(data_tbl, cat_varname, count) {
    bs_tbl <- data_tbl %>%
        sample_n(count) %>%
        group_by_(cat_varname) %>%
        summarise(count = n()) %>%
        mutate(prop = count / sum(count))
    
    return(bs_tbl)
}
```


## Categorical Variables

Dealing with the bootstrap for categorical variables is not obvious to me.

```{r calculate_categorical_bootstrap, echo=TRUE}
for(plot_var in type_list$categorical) {
    subpop_prop_tbl <- subpop_tbl %>%
        group_by_(plot_var) %>%
        summarise(count = n()) %>%
        mutate(prop = count / sum(count)
              ,idx  = 1) %>%
        arrange_(plot_var)

    catprop_lst <- list()
    
    for(i in 1:bootstrap_count) {
        catprop_lst[[i]] <- lifebook_tbl %>%
            generate_bootstrap_props(plot_var, subpop_tbl %>% nrow) %>%
            mutate(idx = i)
    }
    
    bootprops_tbl <- catprop_lst %>% bind_rows()
    rm(catprop_lst)
    
    cat_plot <- ggplot(bootprops_tbl) +
        geom_line(aes_string(x = plot_var, y = 'prop', group = 'idx')
                 ,alpha = 0.1) +
        geom_line(aes_string(x = plot_var, y = 'prop', group = 1)
                 ,data = subpop_prop_tbl
                 ,colour = 'red') +
        expand_limits(y = 0) +
        xlab(plot_var) +
        ylab("Proportion") +
        theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
    
    print(cat_plot)
}
```

## Numeric Variables

We take bootstrap samples of the full dataset, calculate the statistics and compare them to the corresponding statistic in the subpopulation.

```{r calculate_numerical_bootstrap, echo=TRUE}
for(plot_var in type_list$numeric) {
    cat(paste0("Plot Var: ", plot_var, "\n"))

    var_vals <- lifebook_tbl %>% .[[plot_var]]
    var_vals <- var_vals[!is.na(var_vals)]

    subpop_stats_tbl <- var_vals %>%
        calc_bootstrap_stats %>%
        t %>%
        as_data_frame %>%
        gather('variable','value')

    data_boot <- boot(var_vals, calc_bootstrap_stats, R = bootstrap_count)    
    
    boot_tbl        <- data_boot$t %>% as_data_frame
    names(boot_tbl) <- subpop_stats_tbl$variable
    
    bootplot_tbl <- boot_tbl %>%
        mutate(iter = 1:n()) %>%
        gather('variable','value',-iter)

    var_plot <- ggplot(bootplot_tbl) +
        geom_density(aes(x = value)) +
        geom_vline(aes(xintercept = value), data = subpop_stats_tbl, colour = 'red') +
        facet_wrap(~variable, scales = 'free') +
        scale_x_continuous(labels = comma) +
        xlab(plot_var)
        
    print(var_plot)
}
```

## Date/Time Variables

```{r calculate_datetime_bootstrap, echo=TRUE}
for(plot_var in type_list$datetime) {
    cat(paste0("Plot Var: ", plot_var, "\n"))

    var_vals <- lifebook_tbl %>% .[[plot_var]]
    var_vals <- var_vals[!is.na(var_vals)]

    stat_vals <- var_vals %>% calc_bootstrap_stats

    subpop_stats_tbl <- data_frame(variable = names(stat_vals)
                                  ,value    = stat_vals)
    
    data_boot <- boot(var_vals, calc_bootstrap_stats, R = bootstrap_count)    
    
    boot_tbl        <- data_boot$t %>% as_data_frame
    names(boot_tbl) <- subpop_stats_tbl$variable
    
    bootplot_tbl <- boot_tbl %>%
        mutate(iter = 1:n()) %>%
        gather('variable','value',-iter) %>%
        mutate(value = as.Date(value, origin = '1970-01-01'))

    var_plot <- ggplot(bootplot_tbl) +
        geom_histogram(aes(x = value), bins = 50) +
        geom_vline(aes(xintercept = as.numeric(value)), data = subpop_stats_tbl, colour = 'red') +
        facet_wrap(~variable, scales = 'free') +
        scale_x_date(date_labels = '%Y-%m-%d') +
        xlab(plot_var) +
        theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
        
        
    print(var_plot)
}
```

